Ecological Metadata Language interface for R: facilitating synthesis and integration of heterogenous data

### 2014-03-28 inst/doc/pubs/manuscript.Rmd
scratching away at manuscript outline

The future belongs to those who can leverage existing data and provide reusable data.

1. Following metadata standards facilitates reuse
2. data 
3. EML provides an easy way to utilize the Ecological Metadata Language to generate, access, and publish data.

@@

### Implementation and architecture

The `EML` package is built for the Ecological Markup Language schema, [version 2.1.1], using the R software environment and the XML package for parsing and generating XML [Temple Lang 2013, XML].

<!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
<!-- Reference any installation and configuration instructions necessary (e.g. API credentials for figshare) -->

@@

The `EML` package can perform three levels of validation on EML files written from or read into R: validating the XML, validating against the EML 2.1.1 Schema, and performing

Several mechanisms are in place to evaluate the quality of the `EML` package functions themselves.  In addition to the examples from the documentation and the automated package checks provided for the R system, the `EML` package includes a complete test suite using the testthat package [Wickham 2011].

[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html

@@

The `EML` package is available through the CRAN repository under a CC0 license. At the time of writing, the current version is `r packageDescription("EML", fields="Version")`. The current development branch can be found on Github at [https://github.com/ropensci/EML].  See the NEWS file for changes in recent versions. Users are encouraged to submit bug reports or feature requests in the issues log.  Contact ropensci-discuss@googlegroups.com or the package maintainer for help.

@@

OLD OUTLINE
----------------------------------------------------------------------------------------------

It's all metadata.

Share 
Metadata
raw data
formats: file formats, table formats, cell formats 
good null values
easy to combine (context data)
quality control
repository
license

Abstract

1. Recording good metadata is at the heart of effective data synthesis and reuse
2.

Vertically integrated repositories vs metadata driven repositories.

Data synthesis

## Why EML?

But specifying metadata is typically onerous, time-consuming and error prone.

Such definitions are also not machine readable.

Or at least, it doesn't have to be that way.  We just need better tools.

## Why REML?

* The Ecological Metadata Language (EML) has been around for a while now, but has so far seen limited use.

* R is a widely adopted among the community, many researchers already familiar with reading and writing data from R.

* R provides a rich set of data types and constructs that can be mapped into EML (e.g. integers, floats, ordered/nominal factors, character strings)

### Current state of EML

* Generation of EML is limited.  e.g. to LTER centers with informatics support. 
* Applications of existing EML more limited still (?)

* Software to work with EML is limited to tools with low adoption rates across the community (e.g. Kepler).

 Researchers are isolated from the metadata encoding, instead providing plain-text descriptions in MS Word documents.  Allows for low barrier of entry but also makes a barrier to richer encoding.

### Promise of REML

### Getting Started: A minimal example.

Imagine we are working with a `data.frame` in R analgous to Table 1.

<!--Writing a data.frame into EML -->

Writing this table into EML

### Authoring tools

### Parsing tools

### Publishing tools

>>>>

---
layout: preprint
#layout: review, 11pt
title: "Ecological Metadata Language interface for R: facilitating synthesis and integration of heterogenous data"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Karthik Ram 
    affiliation: Berkeley 
  - name: Class-Thido Pfaff
    affiliation: Leipzig
  - name: Matt Jones 
    affiliation: NCEAS 
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA

abstract: | 
  1. Something about the importance of data, data publication, sharing
     and annotation

  2. Something about EML

  3. The EML package provides automated and scriptable tools to make it
     easy to generate and publish data and metadata using state-of-the-art
     technology and publishing standards.

  4. The EML package also makes it easy to consume richly structured data
     and metadata from rapidly growing archives of repositories such as
     KNB and DataONE, including over 20,000 datasets deposited by Long
     Term Ecological Research (LTER) areas over the past 30 years

  5. The EML package facilitates manipulation and integration of
     heterogeneous data sources by providing simple functions to add rich,
     machine-readable annotations to almost any data set.  These
     descriptions can then be used to enable programmatic recombination and
     synthesis of different data sets, allowing users to both take better
     advantage of their own data collections and more easily integrate
     them with others.

---

<!--
The future belongs to those who can leverage existing data and provide reusable data.  
-->

@@

@@

The `EML` package performs severak levels of validation on EML files
written from or read into R: validating the XML, and validating against
the EML schema (based on the declared version), and performing additional
checks through the EML online validation tool for elements that cannot
be enforced by schema validation alone.  Validation serves to improve
consistency in data encoding which facilitates the use of EML across a
wide variety of platforms and software.

Several mechanisms are in place to evaluate the quality of the `EML`
package functions themselves.  In addition to the examples from
the documentation and the automated package checks provided for the
R system, the `EML` package includes a complete test suite using the
`testthat` package [Wickham 2011]. Package development on Github includes
continuous integration through the `travis` platform, in which the full
test suite is run every time the code is changed.  This not only gives
users confidence that the development copy can be successfully installed,
but should assist other developers seeking to contribute to the package
to test that their proposed changes have not negatively impacted any of
the existing code base.

@@

The `EML` package is available through the CRAN repository under
a CC0 license. At the time of writing, the current version is `r
packageDescription("EML", fields="Version")`. The current development
branch can be found on Github at [https://github.com/ropensci/EML].
See the NEWS file for changes in recent versions. Users are encouraged
to submit bug reports or feature requests in the issues log.  Contact
`ropensci-discuss@googlegroups.com` or the package maintainer for help.

@@
